---
title: "SLAM User Guide"
output: rmarkdown::html_vignette
author: Adrian Hordyk <adrian@bluematterscience.com>
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{SLAM_User_Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r initial_stuff, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction 
The `SLAM` package is designed to simulate fishery dynamics and data, and to apply an age-structured assessment model to fishery data for a short-lived semelparous species such as octopus.

`SLAM` is designed with a monthly time-step and to fit to catch-at-weight data. If available, it will also fit to an index of fishing effort. `SLAM` estimates the selectivity pattern, fishing mortality (F), and the spawning potential ratio (SPR) for each month, as well as reference points related to F and SPR.

## Installation
The `SLAM` package can be installed directly from GitHub. A couple of prerequisites are required to install `SLAM` from GitHub. RTools is required because the package uses [TMB](https://cran.r-project.org/package=TMB). The `devtools` package can check if RTools is installed on your system:

```{r install_devtools}
# Load `devtools` if it's installed
chk <- require('devtools')

# install `devtools` if necessary
if (!chk) {
  install.packages('devtools')
  library('devtools')
}
```

```{r, check_rtools}
# Check if RTools is installed
if (!devtools::find_rtools()) {
  stop('RTools needs to be installed')
}
```
If you get an error message after running the code above, you need to install RTools. Instructions for installing RTools for Windows machines are available [here](https://cran.r-project.org/bin/windows/Rtools/rtools40.html). A quick web search will find similar instructions for  installing RTools on OS machines.

### Installing `SLAM`

Once RTools is installed, `SLAM` can be installed with:

```{r, installSLAM, eval=FALSE} 
devtools::install_github('blue-matter/SLAM')
```

# Applying `SLAM` to Fishery Data

In each new R session, the `SLAM` package must first be loaded into memory:
```{r load_library}
library(SLAM)
```

## Importing Fishery Data
Fishery data is loaded into an object of class `Data`. Fishery data can be loaded from a correctly structured CSV file.

The `Import` function can be used to import fishery data from a correctly formatted CSV file.

The `SLAM` package includes a few files with example data. The `Example_Data` function will 
print out the full file paths to these example files:

```{r example}
Example_Data()
```

You can examine the example files to see the specific format of the CSV files containing the example data. Data files must follow this same structure. 

"Data_Template.csv" is an empty data file that can be used as a template for your data.

The data can be imported from the CSV file using the `Import` function:

```{r import}
MyData <- Import(Example_Data()[2])
```

Once you've imported a `Data` object, use the `Check` function to check that everything is structured correctly:

```{r check_data}
Check(MyData)
```

## Structure of the CSV Data File
The CSV files containing the data for the `SLAM` model has to be structured in a specific way. The easiest method for creating your own data file is to copy and edit one of the example data files. 

The sections below describe the structure of each section of the CSV file.

### Meta-Data
The first 7 rows of the CSV file contain the meta-data used for reporting. 

The text in the first column must be exactly: 'Field', 'Fishery', 'Location', 'Species', 'Common_Name', 'Author', and 'Date'.

The first row of the second column must say 'Value'.

Enter the relevant meta-data information for your fishery in the remaining rows. Missing values are accepted.

### Age Schedules

The first column in row 9 must say 'Age Schedules'.
Row 10 must have the text 'Ages', 'Weight_Age_Mean', 'Weight_Age_SD', 'Maturity_at_Age', 'M_at_Age', 'Post_Spawning_Mortality' in columns 1 to 6. 
Enter the corresponding information in the rows below. There must be values for all ages (rows) and all columns.

### CAW Binned Data
Next, enter the catch-at-weight (CAW) data. 
The row with 'CAW Binned Data' in the first column indicates the beginning of the CAW Data.
The row below this must have 'Year', 'Month', and then numeric values for the mid-points of the CAW bins. The CAW bins must increase sequentially with the same increment between each bin. 

The Year and Month columns should increase sequentially by month. Months should be specified as integers. There can not be any missing months. 

Enter the observed weight composition data for each month in the corresponding row. The default is to enter raw observed numbers, such that summing over the weight bins results in the total sample size for that month.

The CAW Data cannot have any missing rows.

### Indices
The row with the text 'Index Data' in the first column indicates the beginning of the index time-series data. 
The row below this must have the text 'Year', 'Month', 'Mean', 'SD', and 'Type'.

The 'Year' and 'Month' values must exactly match those for the CAW data (i.e., it must start and end at the same years and months as the CAW data). 

Enter the mean value and the standard deviation (log-normal) for any available 'Effort', 'Catch', and/or 'Biomass' indices. Enter the corresponding type ('Effort', 'Catch', 'Biomass') in the 'Type' column. 

The data-sets should appear below each other (e.g., first Effort then Catch). 

Missing values should be indicated with NA.

### Parameters 
The final information to enter is parameters for the assessment model. If the parameters are not provided, default values will be used.

The row with text 'Parameters', 'Value', and (optionally) 'Description' indicates the beginning of the parameters section.
The first column below this should have the text: 'Steepness', 'sigmaR', 'sigmaF', and 'sigmaR0'. Enter the corresponding values in the second column.


## Plotting the Data 

Once a `Data` object is complete, the assumed life-history information, the weight composition data, and the indices (if any) can be plotted:


```{r atage}
plot_at_Age(MyData)
```

```{r caw, fig_height=8}
p <- plot_CAW(MyData)
p$p
```

```{r indices}
plot_Indices(MyData)
```

The `plot_CAW` return a list with  `ggplot` object (`p`) and a data.frame containing the information presented in the plots (`df`). 

### Data Report
A HTML Data Report including all these plots can be generated with the `Report` function:

```{r report_data, eval=FALSE}
Report(MyData)
```

This will compile a HTML report with figures and tables summarizing your fishery data, and open the HTML file in your web browser.

See [here](Data_Report.html){target="_blank"} for the example Data Report created above. 

```{r report_data2, include=FALSE}
Report(MyData, dir='../docs/articles', open_file = FALSE)
```

## Running Assessment

Now that the Data is imported and checked, we are ready to run the `SLAM` assessment model. This is done with the `Assess` function:

```{r assess}
MyAssess <- Assess(MyData)
```

### Plotting the Assessment Results

The assessment output can be plotted, e.g.,:

```{r caw2, fig_height=8}
p <- plot_CAW(MyAssess)
p$p
```

```{r indices2}
plot_Indices(MyAssess)
```

```{r spr}
plot_SPR(MyAssess)
```

The `Report` function can also be used to generate an Assessment Report:

```{r report_assessment, eval=FALSE}
Report(MyAssess)
```

The example assessment report can be found [here](Assessment_Report.html){target="_blank"}. 

```{r report_assessment2, include=FALSE}
Report(MyAssess, dir='../docs/articles', open_file = FALSE)
```





