---
title: "SLAM User Guide"
output: rmarkdown::html_vignette
author: Adrian Hordyk <adrian@bluematterscience.com>
date: 
vignette: >
  %\VignetteIndexEntry{SLAM_User_Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction 
The `SLAM` package is designed to simulate fishery dynamics and data, and to apply an age-structured assessment model to simulated or real fishery data, for a short-lived semelparous species such as octopus.

## Installation
The `SLAM` package can be installed directly from GitHub. 

### Prerequisites 
A couple of prerequisites are required to install `SLAM` from GitHub. RTools is required because the package uses [TMB](https://cran.r-project.org/package=TMB). The `devtools` package can check if RTools is installed on your system:

```{r install_devtools}
# Load `devtools` if it's installed
chk <- require('devtools')

# install `devtools` if necessary
if (!chk) {
  install.packages('devtools')
  library('devtools')
}
```

```{r, check_rtools}
# Check if RTools is installed
if (!devtools::find_rtools()) {
  stop('RTools needs to be installed')
}
```
If you get an error message after running the code above, you need to install RTools. Instructions for installing RTools for Windows machines are available [here](https://cran.r-project.org/bin/windows/Rtools/rtools40.html). A quick web search will find similar instructions for  installing RTools on OS machines.

### Installing `SLAM`

Once RTools is installed, `SLAM` can be installed with:

```{r, installSLAM} 
devtools::install_github('blue-matter/SLAM')
```

# Applying `SLAM` to Fishery Data

```{r setup}
library(SLAM)
```

## Importing Data


## Running Assessment

## Examine Results 

# Simulating Fishery Data




## Simulation

```{r Simulate}
Import_LifeHistory()
Simulate()

```
### Simulating Fishery Dynamics 


### Generating Fishery Data 


## Estimation

### Applying Assessment Model to Simulated Data

### Applying Assessment Model to Case Study Data
